FITS format for planetary surfaces: definitions, applications and best practices.

### 2018-02-05 ess_marmo.tex
Update on Overleaf.

Table \ref{table:wcskeys}

\begin{table}
\label{table:wcskeys}
\caption{WCS FITS keywords in the Planetary Surface context}
\centering
\begin{tabular}{l l l l}
\hline
Keyword & Type & Definition & PDS4 \\
\hline
EQUINOX & real & & \\
RADESYS	& string & & \\
CRPIXn & real & & \\
CTYPEn & string & & \\
CRVALn & real & & \\
CDi\_j & real matrix & & \\
PCi\_j & real matrix & & \\
CDELTn & real & & \\
CUNITn & string & & \\
WCSNAME & string & & \\
CNAMEn & string (up to 68 characters) & & \\
\hline
\end{tabular}
\end{table}

@@

\item{\texttt{BLANK}.
This keyword must be used only in integer data.
It contains an integer that specifies the value that is used to
represent pixels that have an undefined physical value.... 
\texttt{PDS4 Special\_Constants?}}

@@

\item{\texttt{DATE-OBS}.
It contains the date of the observation, provided as ISO 8601
string. The value shall be assumed to refer to the start of the observation,
unless another interpretation is clearly explained in the comment field.
It is assumed to be expressed in UTC but other systems are accepted as long as they
are specified in the ISO string. 
PDS4 provides a \texttt{Time\_Coordinates} class with attributes
\textit{start\_date\_time} and \texttt{stop\_date\_time} providing start and
stop date and time appropriate to the product being labeled, always in UT
coordinates.
EPN-TAP also strongly recommends to fill in the \texttt{time\_min} and
\texttt{time\_max} fields, always in UTC and formatted in Julian Days.}
% * <chiara.marmo@u-psud.fr> 2018-02-05T10:20:13.335Z:
% 
% > Julian Days
% From https://voparis-confluence.obspm.fr/display/VES/EPN-TAP+v2+parameter+description I see that ISO format is still in the race.... Maybe it would be better to homogenize PDS4 EPN-TAP and FITS?
% 
% ^.

\item{\texttt{INSTRUME}.}
\item{\texttt{TELESCOP}}
\item{\texttt{OBJECT}}
\item{\texttt{ORIGIN}}
\item{\texttt{AUTHOR}}
\item{\texttt{REFERENC}}

@@

\subsection{Best Practices and Examples}

\section{Inter-operable software developments}
\label{sec:softdev}

\section{Discussion and Perspectives}
\label{sec:disc}

>>>>

Once reprojected in one of the standard projections listed in Table \ref{table:proj}
Planetary images or maps can be described using the WCS scheme
\citep{greisencalaII,calagreisenI}.
In WCS context pixels are first translated in an intermediate pixel coordinate system
via linear transformations, then mapped on the grid specified by the \texttt{CTYPE}
keyword, containing the projection code and the coordinate system.
The coordinate system must be defined with respect to a specific inertial reference frame.

Multiple coordinate system descriptions are allowed in the same header: this is a valuable
property for planetary surface imagery, as angular ('deg') and linear ('m') description of
the surface are often needed at the same time.
Further below 3D keyword will be proposed for FITS metadata, in order to standardize
the conversion between angular and linear coordinates: alternate description will still
be important in our framework as standard libraries (e.g.
wcslib\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/} or
astropy\footnote{http://www.astropy.org/} \citet{astropy}) and some FITS visualization
software (e.g. ds9\footnote{http://ds9.si.edu/}) already implement them.

Table \ref{table:wcskeys} describes FITS WCS keywords and their recommended use in
planetary surface context.

\begin{sidewaystable}
\label{table:wcskeys}
\caption{WCS FITS keywords in the Planetary Surface context}
\centering
\begin{tabular}{l l l l l}
\hline
Keyword & Type & Definition & PDS4 & EPN-TAP \\
\hline
RADESYS	& string & Name of the inertial reference frame with respect & & \\
& & to which the coordinate system is defined & & \\
& & (defaulted to \texttt{'ICRS'}) & & \\
EQUINOX & real & The value field contains the equinox in years for the celestial & & \\
& & coordinate system in which positions are expressed & & \\
& & (defaulted to \texttt{'J2000'}) & & \\
WCSAXES & integer & Number of axes in the WCS description & & \\
& & The value of WCSAXES may exceed the number of raster pixel axes & & \\
WCSNAME & & & & \\
CRPIXn & real & & \\
CTYPEn & string & & \\
CRVALn & real & & \\
CDi\_j & real matrix & & \\
PCi\_j & real matrix & & \\
CDELTn & real & & \\
CUNITn & string & & \\
CNAMEn & string (up to 68 characters) & & \\
\hline
\end{tabular}
\end{sidewaystable}

@@

\item{\texttt{BLANK}.
This keyword must be used only in integer data.
It contains an integer that specifies the value that is used to
represent pixels that have an undefined physical value.
On floating point data the IEEE NaN must be used to represent undefined values.  
\texttt{PDS4 Special\_Constants?}}

@@

\item{\texttt{DATE-OBS}.
It contains the date of the observation, provided as ISO 8601
string. The value shall be assumed to refer to the start of the observation,
unless another interpretation is clearly explained in the comment field.
It is assumed to be expressed in UTC but other systems are accepted as long as they
are specified in the ISO string. 
PDS4 provides a \texttt{Time\_Coordinates} class with attributes
\textit{start\_date\_time} and \texttt{stop\_date\_time} providing start and
stop date and time appropriate to the product being labeled, always in UT
coordinates.
The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP scheme,
they must always be formatted in UTC and Julian Days.}
% * <chiara.marmo@u-psud.fr> 2018-02-05T10:20:13.335Z:
% 
% > Julian Days
% From https://voparis-confluence.obspm.fr/display/VES/EPN-TAP+v2+parameter+description I see that ISO format is still in the race.... Maybe it would be better to homogenize PDS4 EPN-TAP and FITS?
% 
% ^.

\item{\texttt{INSTRUME}.
It contains a character string identifying the instrument used to acquire the data.
PDS4 provides an \texttt{Instrument} class which attribute \texttt{Name} ideally corresponds
to this FITS keywords.
EPN-TAP provides with an \texttt{instrument\_name} parameter.}

\item{\texttt{TELESCOP}.
It contains a character string identifying the telescope used to acquire the data.
The PDS4 \texttt{Instrument\_Host} class provides the description of the physical object
mounting the instrument. Its \texttt{Name} attribute should correspond to the FITS keyword.
PDS4 also provides for the definition of a \texttt{Telescope} class, in case of
ground-based observations.
In the EPN-TAP scheme this will defined the \texttt{instrument\_host\_name} parameter.
\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Spacecraft}}

\item{\texttt{OBJECT}.
It contains a character string giving a name for the observed object\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Planets\%20and\%20Satellites}.
The PDS4 \texttt{Target} class contains the \texttt{name} attribute corresponding to this
FITS keyword.
EPN-TAP standard also provides with a \texttt{target\_name} parameter.
Notice that PDS4 and EPN-TAP also ask for a \texttt{target\_type} attribute, which must be
specified from an enumerated list.}

%
%\item{\texttt{ORIGIN}.
%It contains string identifying the organization or institution responsible for
%creating the FITS file.
%}
%
%\item{\texttt{AUTHOR}}

\item{\texttt{REFERENC}.
It contains a character string citing a reference where the data are published.
It is recommended that either the Astrophysics Data System bibliographic databases\footnote{http://adswww.harvard.edu/}
bibliographic identifier or the Digital Object Identifier\footnote{http://doi.org}
be used.
PDS4 \texttt{External\_Reference} class 
EPN-TAP \texttt{bib\_reference} has exactly the same meaning and use than the FITS keyword.}

@@

\subsection{Examples and Best Practices}
\subsubsection{Global maps}
\subsubsection{HiRISE mosaics}
\subsubsection{OMEGA spectra and geometry data}
\subsubsection{Akatsuki imaging data}
\section{Inter-operable software developments}
\label{sec:softdev}

\section{Discussion and Perspectives}
\label{sec:disc}
FITS discussion in astronomy\citep{THOMAS2015133} 
let's astronomy learn from planetary sciences too: accent must be put on interoperability
.... blabla
Planetary surface is lacking open and modular software for massive processing and can benefit
of astronomy experience. In addition to that WCS projection scheme is simpler than the
historical one used in planetary applications. 
Archiving flexibility and data model are today approached by PDS4 and EPN-TAP efforts.
Reducing data format conversion during the processing chain but
allowing easy conversion for final products following the needed application.
